---
title: "Basic R Programming"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(learnr)
library(tidyverse)
library(here)
library(knitr)
opts_chunk$set(echo = TRUE)
tutorial_options(exercise.timelimit = 10)
```

```{r echo=F, message=F} 
# actual data read-in. Not showing in tutorial because don't want to confuse them with file path
survey_data_2 <- read_csv(here('Basic_R_Programming','data','survey_data_2.csv'))
```

## **Introduction**
  
In this tutorial, we will go over the basics of programming with `R` and how we can use `R` to work with data.
  
## `R` as a calculator
  
At its most basic level, `R` can be used to perform calculations just like your regular handheld calculator. Any arithmetic/mathematical operation can be done in `R`. 
  
```{r}
2+2
5-1
4*4
10/5
```
  
### Exercise - Use `R` as a calculator
  
Try out using `R` as a calculator! 
  
**OPTIONAL** - ADVANCED EXERCISE: Use `R` as a calculator and compute the number of minutes in a year.
  
```{r calculator, exercise=TRUE, exercise.lines=10}

```
  
```{r calculator-solution}
# any arithmetic will do
5+3
8-1
7*2
24/3
# OPTIONAL ADVANCED ANSWER
365*60*24
```

## Basic `R` Syntax 
  
Now that we know a little bit about using `R` as a calculator, let's get into basic `R` syntax. 
  
`R` works by creating objects - pieces of data stored in the *environment*. We can create an object `x` that contains a single piece of data - the number `5`.
  
```{r}
x <- 5
```
  
When we type the object, `x`, and run it, `R` will print out the contents.
```{r}
x
```
  
`x` is an example of a **vector**, an object containing values of a single **type**. Here, the type is **numeric**, because 5 is a number. Furthermore, `x` contains only one value. However, we can use the `c()` command to create a vector containing multiple values. All we have to do is use `c()` and separate the values with a comma. Here's an example:
  
```{r}
vector_1 <- c(1,2,3,4,5)
vector_1
```
  
Again, when we type the object name, `R` will print its contents.  
  
We can perform arithmetic operations on this vector, as well.

```{r}
vector_1 + 1
```
  
Because `R` is **vectorized**, this operation was performed on *each element* of `vector_1`. That means that we add 1 to the first value in `vector_1`, `1`, and then do so for the second value, `2`, and so on. 
