---
title: "Basic R Programming"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(learnr)
library(tidyverse)
library(here)
library(knitr)
opts_chunk$set(echo = TRUE)
tutorial_options(exercise.timelimit = 10)
```

```{r echo=F, message=F} 
# actual data read-in. Not showing in tutorial because don't want to confuse them with file path
survey_data_2 <- read_csv(here('Basic_R_Programming','data','survey_data_2.csv'))
```

## **Introduction**
  
In this tutorial, we will go over the basics of programming with `R` and how we can use `R` to work with data.
  
## `R` as a calculator
  
At its most basic level, `R` can be used to perform calculations just like your regular handheld calculator. Any arithmetic/mathematical operation can be done in `R`. 
  
```{r}
2+2
5-1
4*4
10/5
```
  
### Exercise - Use `R` as a calculator
  
Try out using `R` as a calculator! 
  
**OPTIONAL** - ADVANCED EXERCISE: Use `R` as a calculator and compute the number of minutes in a year.
  
```{r calculator, exercise=TRUE, exercise.lines=10}

```
  
```{r calculator-solution}
# any arithmetic will do
5+3
8-1
7*2
24/3
# OPTIONAL ADVANCED ANSWER
365*60*24
```

## Basic `R` Syntax 
  
Now that we know a little bit
